---
title: "QuickSort Performance Analysis - Sequential vs Parallel"
author: "Lamine"
date: "2025-10-08"
output: html_document
---

## Confidence Interval Analysis

### What I Did
I improved the original experiment by:

- Increasing repetitions from 5 to 20 (for better statistical confidence)
- Testing more array sizes: 50k, 100k, 200k, 300k, 400k, 500k, 600k, 700k, 800k, 900k, 1M
- Calculating 95% confidence intervals using the t-distribution

### Understanding Confidence Intervals
A confidence interval shows the range where the true mean is likely to be.

Formula I used:
```
CI = mean ± t × (sd / √n)
```

Where:

- mean = average of 20 measurements
- sd = standard deviation
- n = 20 repetitions
- t ≈ 2.09 (from t-table for 95% CI with 19 degrees of freedom)

The error bars in my plot show these confidence intervals.

## Data Analysis

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
```

### Loading Data

```{r load-data}
df <- read.csv("data/sama_2014-10-13/measurements_03:47.csv", 
               header = TRUE, 
               strip.white=TRUE)
```

### Calculating Confidence Intervals

```{r calculate-stats}
# Filter for Sequential and Parallel types
df_filtered <- subset(df, Type =="Sequential" | Type == "Parallel")

# Calculate statistics with 95% confidence intervals
stats <- aggregate(Time ~ Size + Type, data=df_filtered,
                    FUN = function(x){
                        n <- length(x)
                        mean_x <- mean(x)
                        sd_x <- sd(x)
                        se <- sd_x/ sqrt(n)
                        
                        t_val <- qt(0.975, df=n-1)
                        c(mean=mean_x,
                           lower=mean_x - t_val*se,
                           upper=mean_x + t_val*se)
                    })

stats <- data.frame(
    Size = stats$Size,
    Type = stats$Type,
    mean = stats$Time[,1],
    lower = stats$Time[,2],
    upper = stats$Time[,3]
)

print("Mean Times with 95% Confidence Intervals:")
print(stats)
```

### Visualization

```{r plot, fig.width=10, fig.height=6}
ggplot(stats, aes(x=Size, y=mean, color=Type))+
    geom_line() +
    geom_point(size=3) +
    geom_errorbar(aes(ymin=lower, ymax=upper), width=20000) +
    labs(title="Sequential vs Parallel QuickSort with 95% Confidence Intervals",
         x="Array Size",
         y="Time (seconds)") +
    theme_minimal()

ggsave("simple_ci_plot.png", width=10, height=6)
```

## Key Findings

### 1. Thread Overhead is Significant
At small array sizes (50k-100k), the parallel version is SLOWER:

- 50,000 elements: Parallel ≈0.020s vs Sequential ≈0.003s 
- 100,000 elements: Parallel ≈0.048s vs Sequential ≈0.023s 

### 2. Parallel Becomes Faster at Large Sizes
At 1,000,000 elements: Parallel ≈0.162s vs Sequential ≈0.235s

### 3. Crossover Point
Based on my plot, the crossover (where parallel becomes faster) 
happens somewhere between 250,000 and 500,000 elements.
I would need more data points in this range to find the exact crossover.

## What I Learned

1. **Statistical Rigor Matters**: 
   With only 5 repetitions (original), confidence intervals would be much wider.
   20 repetitions gave me tighter bounds and more confidence in my results.

2. **Thread Overhead is Real**:
   Parallelization is NOT always faster! For small problems, the overhead
   of creating and managing threads can make it slower.

3. **Plotting and playing with data is fun**:
   Tinkering with ggplot and using various elements to calculate different things gave me
   a lot of insights to where to direct the experiment.

4. **Testing with fine grained samples**:
   I should add more points (like 200k, 300k, 400k, 500k) to find the exact crossover point.

## Limitations & Future Work

- Missing data points between 100k-1M make crossover estimation imprecise
- THREAD_LEVEL is fixed at 10 - I should test with different values
- Only tested on one machine (Linux)

## My Contribution to the Project

- Designed and ran improved experiment (20 reps, more sizes)
- Implemented confidence interval calculation in R
- Created visualization showing performance comparison
- Identified key performance characteristics of the parallel implementation

