---
title: "QuickSort Performance Analysis - Sequential vs Parallel"
author: "Lamine"
date: "2025-10-08"
output: html_document
---

## Experiment Overview

I improved the original experiment by:

- Increasing repetitions from 5 to 20 (for better statistical confidence)
- Testing more array sizes: 50k, 100k, 200k, 300k, 400k, 500k, 600k, 700k, 800k, 900k, 1M
- Calculating 95% confidence intervals using bootstrap resampling

## Data Analysis

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(boot)
```

### Loading Data

```{r load-data}
# Loading the data with all array sizes (50k-1M)
df <- read.csv("data/linuxmachine_2025-10-08/measurements_22:54.csv", 
               header = TRUE, 
               strip.white=TRUE)
```

### Calculating Confidence Intervals

```{r calculate-stats}
# Filter for Sequential and Parallel types
df_filtered <- subset(df, Type =="Sequential" | Type == "Parallel")

# Bootstrap function to calculate mean
boot_mean <- function(data, indices) {
    return(mean(data[indices]))
}

# Calculate statistics with 95% bootstrap confidence intervals
stats <- aggregate(Time ~ Size + Type, data=df_filtered,
                    FUN = function(x){
                        n <- length(x)
                        mean_x <- mean(x)
                        
                        # Bootstrap confidence interval
                        boot_results <- boot(x, boot_mean, R=10000)
                        ci <- boot.ci(boot_results, type="perc")
                        
                        c(mean=mean_x,
                          lower=ci$percent[4],
                          upper=ci$percent[5])
                    })

stats <- data.frame(
    Size = stats$Size,
    Type = stats$Type,
    mean = stats$Time[,1],
    lower = stats$Time[,2],
    upper = stats$Time[,3]
)

print("Mean Times with 95% Confidence Intervals:")
print(stats)
```

### Visualization

```{r plot, fig.width=10, fig.height=6}
ggplot(stats, aes(x=Size, y=mean, color=Type))+
    geom_line() +
    geom_point(size=3) +
    geom_errorbar(aes(ymin=lower, ymax=upper), width=20000) +
    labs(title="Sequential vs Parallel QuickSort with 95% Confidence Intervals",
         x="Array Size",
         y="Time (seconds)") +
    theme_minimal() +
    scale_x_continuous(labels = scales::comma)

ggsave("simple_ci_plot.png", width=10, height=6)
```

## Key Findings

### 1. Thread Overhead at Small Sizes
At 50,000 elements, Parallel QuickSort (~0.052 seconds) was substantially slower than Sequential QuickSort (~0.010 seconds), confirming significant thread overhead at smaller input sizes.

### 2. Crossover Point Identified at 700,000 Elements
The data confirms a precise crossover point:
- **At 600,000 elements**: Sequential (0.1493s) was faster than Parallel (0.1594s)
- **At 700,000 elements**: Parallel (0.1779s) became faster than Sequential (0.1791s) ✓
- **At 800,000 elements**: Parallel (0.1994s) maintained advantage over Sequential (0.2127s)

### 3. Performance Advantage Beyond Crossover
For array sizes above 700,000 elements, parallel implementation consistently outperforms sequential, with the gap widening as array size increases.

## Lessons Learned

1. **Statistical Rigor Matters**: 20 repetitions with bootstrap confidence intervals gave precise measurements to identify the exact crossover point at 700,000 elements.

2. **Bootstrap for Performance Data**: Bootstrap CI is more appropriate than t-distribution because performance measurements are often not normally distributed.

3. **Thread Overhead is Real**: Parallelization is NOT always faster! For arrays smaller than 700,000 elements, thread overhead dominates.

4. **Practical Optimization**: Parallelization should only be applied to array sizes above 700,000 elements for this implementation.

## Future Work

- Investigate specific factors contributing to thread overhead (thread creation, synchronization costs) to potentially reduce the crossover point
- Test with different thread levels to optimize performance
- Validate findings on different hardware configurations


